\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-12-23 mSystems/rev/mSys_MAIN_rev.tex
new study in the text, several minor modifications
Minor modifications and update Fig2 source scripts&data
Changes to be committed:
	modified:   ../README.md
	modified:   figs/Fig2/ellipse.m
	modified:   figs/Fig2/ellipsedata.txt
	modified:   rev/mSys_MAIN_rev.pdf
	modified:   rev/mSys_MAIN_rev.tex
	modified:   rev/mSys_discussion_rev.tex
Update supplementary, Figure 2 and various minor enhancements
Changes to be committed:
	modified:   rev/figs/Fig2.eps
	new file:   rev/figs/Fig2/Fig2.pdf
	new file:   rev/figs/Fig2/Fig2.svg
	new file:   rev/figs/Fig2/README
	new file:   rev/figs/Fig2/ellipse.

% release codenumber (update it here!)
\newcommand\releasecode{REVISION draft A}

@@

%% Results
\bibitem{fs} {\bf Eisler Z, Bartos I, Kertész J.} 2008. Fluctuation scaling in complex systems: Taylor’s law and beyond1. Adv Phys {\bf 57}:89–142.
% -> main studies
\bibitem{moving} {\bf Caporaso JG, Lauber CL, Costello EK, Berg-Lyons D, Gonzalez A, Stombaugh J, Knights D, Gajer P, Ravel J, Fierer N, Gordon JI, Knight R.} 2011. Moving pictures of the human microbiome. Genome Biol {\bf 12}:R50.
\bibitem{antibiotic} {\bf Dethlefsen L, Relman DA.} 2011. Incomplete recovery and individualized responses of the human distal gut microbiota to repeated antibiotic perturbation. Proc Natl Acad Sci {\bf108}:4554–61.
\bibitem{LEA} {\bf Faith JJ, Guruge JL, Charbonneau M, Subramanian S, Seedorf H, Goodman AL, Clemente JC, Knight R, Heath AC, Leibel RL, Rosenbaum M, Gordon JI.} 2013. The long-term stability of the human gut microbiota. Science {\bf 341}:1237439.
\bibitem{kwashiorkor} {\bf Smith MI, Yatsunenko T, Manary MJ, Trehan I, Mkakosya R, Cheng J, Kau AL, Rich SS, Concannon P, Mychaleckyj JC, Liu J, Houpt E, Li J V, Holmes E, Nicholson J, Knights D, Ursell LK, Knight R, Gordon JI.} 2013. Gut microbiomes of Malawian twin pairs discordant for kwashiorkor. Science {\bf 339}:548–54.
\bibitem{diet} {\bf David LA, Maurice CF, Carmody RN, Gootenberg DB, Button JE, Wolfe BE, Ling A V, Devlin AS, Varma Y, Fischbach MA, Biddinger SB, Dutton RJ, Turnbaugh PJ.} 2014. Diet rapidly and reproducibly alters the human gut microbiome. Nature {\bf 505}:559–63.
% <- END main studies
\bibitem{stat} {\bf Jørgensen B, Martinez JR, Tsao M.} 1994. Asymptotic behaviour of the variance function. Scand J Stat {\bf 21}:223–243.
\bibitem{convergence1} {\bf Fronczak A, Fronczak P.} 2010. Origins of Taylor’s power law for fluctuation scaling in complex systems. Phys Rev E {\bf 81}:066112.
\bibitem{convergence2} {\bf Kendal, W.S., Jorgensen,B.} Taylor's power law and fluctuation scaling explained by a central-limit-like convergence.  Phys. Rev. E {\bf 83} 066115.
\bibitem{convergence3} {\bf Kendal WS, Jørgensen B.} 2011. Tweedie convergence: A mathematical basis for Taylor’s power law, 1/f noise, and multifractality. Phys Rev E {\bf 84}:066120.

@@

\caption{Taylor parameters for persons diagnosed with irritable bowel syndrome (IBS)\cite{IBS}. Healthy individuals sampled in this study are characterized by $\bar{V} = 0.134 \pm 0.009, \bar{\beta} = 0.691 \pm 0.025$.}

@@

\caption{Taylor parameters for individuals with different degrees of overweight and obesity\cite{LEA}. Healthy people in this study, whom were not obese, are characterized by $\bar{V} = 0.19 \pm 0.06, \bar{\beta} = 0.806 \pm 0.034$.} 
\end{table}

>>>>

% release codenumber (update it here!)
\newcommand\releasecode{REVISION draft C}

@@

%% Results
\bibitem{fs} {\bf Eisler Z, Bartos I, Kertész J.} 2008. Fluctuation scaling in complex systems: Taylor’s law and beyond1. Adv Phys {\bf 57}:89–142.
% -> main studies
\bibitem{moving} {\bf Caporaso JG, Lauber CL, Costello EK, Berg-Lyons D, Gonzalez A, Stombaugh J, Knights D, Gajer P, Ravel J, Fierer N, Gordon JI, Knight R.} 2011. Moving pictures of the human microbiome. Genome Biol {\bf 12}:R50.
\bibitem{antibiotic} {\bf Dethlefsen L, Relman DA.} 2011. Incomplete recovery and individualized responses of the human distal gut microbiota to repeated antibiotic perturbation. Proc Natl Acad Sci {\bf108}:4554–61.
\bibitem{LEA} {\bf Faith JJ, Guruge JL, Charbonneau M, Subramanian S, Seedorf H, Goodman AL, Clemente JC, Knight R, Heath AC, Leibel RL, Rosenbaum M, Gordon JI.} 2013. The long-term stability of the human gut microbiota. Science {\bf 341}:1237439.
\bibitem{kwashiorkor} {\bf Smith MI, Yatsunenko T, Manary MJ, Trehan I, Mkakosya R, Cheng J, Kau AL, Rich SS, Concannon P, Mychaleckyj JC, Liu J, Houpt E, Li J V, Holmes E, Nicholson J, Knights D, Ursell LK, Knight R, Gordon JI.} 2013. Gut microbiomes of Malawian twin pairs discordant for kwashiorkor. Science {\bf 339}:548–54.
\bibitem{diet} {\bf David LA, Maurice CF, Carmody RN, Gootenberg DB, Button JE, Wolfe BE, Ling A V, Devlin AS, Varma Y, Fischbach MA, Biddinger SB, Dutton RJ, Turnbaugh PJ.} 2014. Diet rapidly and reproducibly alters the human gut microbiome. Nature {\bf 505}:559–63.
\bibitem{hostlife} {\bf David LA, Materna AC, Friedman J, Campos-Baptista MI, Blackburn MC, Perrotta A, Erdman SE, Alm EJ.} 2014. Host lifestyle affects human microbiota on daily timescales. Genome Biol {\bf 15}:R89.
% <- END main studies
\bibitem{stat} {\bf Jørgensen B, Martinez JR, Tsao M.} 1994. Asymptotic behaviour of the variance function. Scand J Stat {\bf 21}:223–243.
\bibitem{convergence1} {\bf Fronczak A, Fronczak P.} 2010. Origins of Taylor’s power law for fluctuation scaling in complex systems. Phys Rev E {\bf 81}:066112.
\bibitem{convergence2} {\bf Kendal, W.S., Jorgensen,B.} Taylor's power law and fluctuation scaling explained by a central-limit-like convergence.  Phys. Rev. E {\bf 83} 066115.
\bibitem{convergence3} {\bf Kendal WS, Jørgensen B.} 2011. Tweedie convergence: A mathematical basis for Taylor’s power law, 1/f noise, and multifractality. Phys Rev E {\bf 84}:066120.

@@

\caption{Taylor parameters for persons diagnosed with irritable bowel syndrome (IBS)\cite{IBS}. Healthy individuals sampled in this study are characterized by $\bar{V} = 0.135 \pm 0.010, \bar{\beta} = 0.692 \pm 0.024$.}

@@

\caption{Taylor parameters for individuals with different degrees of overweight and obesity\cite{LEA}. Healthy people in this study, whom were not obese, are characterized by $\bar{V} = 0.19 \pm 0.06, \bar{\beta} = 0.806 \pm 0.034$.}

\caption{Taylor parameters for special intervals concerning gut microbiota in the host lifestyle study\cite{hostlife}. The healthy and quotidian periods are characterized by $\bar{V} = 0.25 \pm 0.09, \bar{\beta} = 0.777 \pm 0.025$.} 
\end{table}

### 2016-12-23 mSystems/rev/mSys_discussion_rev.tex
new study in the text, several minor modifications
Minor modifications and update Fig2 source scripts&data
Changes to be committed:
	modified:   ../README.md
	modified:   figs/Fig2/ellipse.m
	modified:   figs/Fig2/ellipsedata.txt
	modified:   rev/mSys_MAIN_rev.pdf
	modified:   rev/mSys_MAIN_rev.tex
	modified:   rev/mSys_discussion_rev.tex
Update supplementary, Figure 2 and various minor enhancements
Changes to be committed:
	modified:   rev/figs/Fig2.eps
	new file:   rev/figs/Fig2/Fig2.pdf
	new file:   rev/figs/Fig2/Fig2.svg
	new file:   rev/figs/Fig2/README
	new file:   rev/figs/Fig2/ellipse.

By solving the Langevin differential equation, we can obtain a phase diagram where each microbiota sample can be placed according to its fitness and variability in one of two phases according to the ranking stability of the system. As we can see in the phase-space in Figure \ref{fig:main3}, we are showing three different conditions that could happen. First, we can have a healthy microbiota which could have some fluctuations as showed by one of the subjects of Caporasso et al study \cite{moving}. Because the fitness of this cases will be high enough, the temporal variability will not place the microbiota in the unstable phase of the diagram. Second, we have a subject from the study of Dethlefsen and Relman \cite{antibiotic} which is perturbed twice by an antibiotic intake. His microbiota is altered enough to lose its stability, and hence be placed in the unstable part. So located, it is more sensitive to potential perturbations as, for example, opportunist infections. Third and last condition, the subject is already in the unstable phase due to some healthy issue as IBS. This can be observed in one of the patients from Durban et al study \cite{IBS}. In addition, it was shown that this subject improved its healthy status in the time when the experiment was done, implying that his microbiota also recovered the lost stability.

>>>>

By solving the Langevin differential equation, we can obtain a phase diagram where each microbiota sample can be placed according to its fitness and variability in one of two phases according to the ranking stability of the system. As we can see in the phase-space in Figure \ref{fig:main3}, we are showing three different conditions that could happen. First, we can have a healthy microbiota which could have some fluctuations as showed by one of the subjects of Caporasso et al study \cite{moving}. Because the fitness of this cases will be high enough, the temporal variability will not place the microbiota in the unstable phase of the diagram. Second, we have a subject from the study of Dethlefsen and Relman \cite{antibiotic} which is perturbed twice by an antibiotic intake. His microbiota is altered enough to lose its stability, and hence be placed in the unstable part. So located, it is more sensitive to potential perturbations as, for example, opportunist infections. Third and last condition, the subject is already in the unstable phase due to some healthy issue as IBS. This can be observed in one of the patients from Durban et al study \cite{IBS}. In addition, it was shown that this subject improved its healthy status in the time when the experiment was done, implying that his microbiota also recovered the lost stability. It is interesting to notice that in the subject from the study of David et al. \cite{hostlife}, who suffered a Salmonella infection during the experiment, we can observe a huge shift in the variability and a final recovery from the perturbed state (see \ref{supfig:HLS_xWSummary}).

\begin{supfig}
	\centering
	%\vspace*{-10mm} % Corrects overbox of the figures
	\includegraphics[width=0.99\textwidth]{figs/supfig_HLS_xWSummary.eps}
	\caption{Taylor's law parameter space for intervals concerning gut microbiota in the host lifestyle study\cite{hostlife}.}
	\label{supfig:HLS_xWSummary}
\end{supfig}

### 2016-12-23 mSystems/rev/mSys_suppl_rev.tex
Update supplementary, Figure 2 and various minor enhancements
Changes to be committed:
	modified:   rev/figs/Fig2.eps
	new file:   rev/figs/Fig2/Fig2.pdf
	new file:   rev/figs/Fig2/Fig2.svg
	new file:   rev/figs/Fig2/README
	new file:   rev/figs/Fig2/ellipse.m
	new file:   rev/figs/Fig2/ellipsedata.txt
	new file:   rev/figs/Fig2/errorbarxy.m
	new file:   rev/figs/supfig_HLS_xWSummary.eps
	new file:   rev/figs/supfig_HLS_xWSummary.pdf
	modified:   rev/mSys_MAIN_rev.pdf
	modified:   rev/mSys_MAIN_rev.tex
	modified:   rev/mSys_discussion_rev.tex
	new file:   rev/mSys_suppl_rev.pdf
	modified:   re

% release codenumber (update it here!)
\newcommand\releasecode{final}

@@

\begin{table}
\caption{Taylor parameters for persons diagnosed with irritable bowel syndrome (IBS)\cite{IBS}. Healthy individuals sampled in this study are characterized by $\bar{V} = 0.134 \pm 0.009, \bar{\beta} = 0.691 \pm 0.025$.}

  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		IBS&$0.204 \pm 0.034$&$0.739 \pm 0.029$&$0.916$&&$7.6 \pm 3.7$&$1.9 \pm 1.2$\\
		IBS&$0.35 \pm 0.05$&$0.793 \pm 0.023$&$0.935$&&$23.1 \pm 5.9$&$4.0 \pm 0.9$\\
	     \hline
	     \hline
    \end{tabular}
  \end{center}
  \label{tab:IBS}
\end{table}

@@

\begin{thebibliography}{9}
\bibitem{diet} {\bf David LA, Maurice CF, Carmody RN, Gootenberg DB, Button JE, Wolfe BE, Ling A V, Devlin AS, Varma Y, Fischbach MA, Biddinger SB, Dutton RJ, Turnbaugh PJ.} 2014. Diet rapidly and reproducibly alters the human gut microbiome. Nature {\bf 505}:559–63.
\bibitem{antibiotic} {\bf Dethlefsen L, Relman DA.} 2011. Incomplete recovery and individualized responses of the human distal gut microbiota to repeated antibiotic perturbation. Proc Natl Acad Sci {\bf108}:4554–61.
\bibitem{IBS} {\bf Durbán A, Abellán JJ, Jiménez-Hernández N, Artacho A, Garrigues V, Ortiz V, Ponce J, Latorre A, Moya A.} 2013. Instability of the faecal microbiota in diarrhoea-predominant irritable bowel syndrome. FEMS Microbiol Ecol {\bf 86}:581–589.
\bibitem{kwashiorkor} {\bf Smith MI, Yatsunenko T, Manary MJ, Trehan I, Mkakosya R, Cheng J, Kau AL, Rich SS, Concannon P, Mychaleckyj JC, Liu J, Houpt E, Li J V, Holmes E, Nicholson J, Knights D, Ursell LK, Knight R, Gordon JI.} 2013. Gut microbiomes of Malawian twin pairs discordant for kwashiorkor. Science {\bf 339}:548–54.
\bibitem{LEA} {\bf Faith JJ, Guruge JL, Charbonneau M, Subramanian S, Seedorf H, Goodman AL, Clemente JC, Knight R, Heath AC, Leibel RL, Rosenbaum M, Gordon JI.} 2013. The long-term stability of the human gut microbiota. Science {\bf 341}:1237439.
\end{thebibliography}
\end{document}

>>>>

% release codenumber (update it here!)
\newcommand\releasecode{REVISION draft C}

@@

\begin{table}
\caption{Taylor parameters for persons diagnosed with irritable bowel syndrome (IBS)\cite{IBS}. Healthy individuals sampled in this study are characterized by $\bar{V} = 0.135 \pm 0.010, \bar{\beta} = 0.692 \pm 0.024$.}

  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		IBS (minor)&$0.205 \pm 0.034$&$0.740 \pm 0.029$&$0.917$&&$6.9 \pm 3.3$&$2.0 \pm 1.2$\\
		IBS (severe)&$0.35 \pm 0.06$&$0.793 \pm 0.025$&$0.934$&&$21 \pm 6$&$4.2 \pm 1.0$\\
	     \hline
	     \hline
    \end{tabular}
  \end{center}
  \label{tab:IBS}
\end{table}

@@

\begin{table}
\caption{Taylor parameters for special intervals concerning gut microbiota in the host lifestyle study\cite{hostlife}. The healthy and quotidian periods are characterized by $\bar{V} = 0.25 \pm 0.09, \bar{\beta} = 0.777 \pm 0.025$.}

  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		HLS (abroad)&$0.51 \pm 0.06$&$0.820 \pm 0.012$&$0.928$&&$2.8 \pm 0.6$&$1.7 \pm 0.5$\\
		HLS (infection)&$0.49 \pm 0.08$&$0.828 \pm 0.018$&$0.923$&&$2.6 \pm 0.9$&$2.0 \pm 0.7$\\
		HLS (after infection)&$0.36 \pm 0.05$&$0.776 \pm 0.015$&$0.922$&&$1.1 \pm 0.6$&$-0.0 \pm 0.6$\\
	     \hline
	     \hline
    \end{tabular}
  \end{center}
  \label{tab:IBS}
\end{table}

\begin{thebibliography}{9}
\bibitem{diet} {\bf David LA, Maurice CF, Carmody RN, Gootenberg DB, Button JE, Wolfe BE, Ling A V, Devlin AS, Varma Y, Fischbach MA, Biddinger SB, Dutton RJ, Turnbaugh PJ.} 2014. Diet rapidly and reproducibly alters the human gut microbiome. Nature {\bf 505}:559–63.
\bibitem{antibiotic} {\bf Dethlefsen L, Relman DA.} 2011. Incomplete recovery and individualized responses of the human distal gut microbiota to repeated antibiotic perturbation. Proc Natl Acad Sci {\bf108}:4554–61.
\bibitem{IBS} {\bf Durbán A, Abellán JJ, Jiménez-Hernández N, Artacho A, Garrigues V, Ortiz V, Ponce J, Latorre A, Moya A.} 2013. Instability of the faecal microbiota in diarrhoea-predominant irritable bowel syndrome. FEMS Microbiol Ecol {\bf 86}:581–589.
\bibitem{kwashiorkor} {\bf Smith MI, Yatsunenko T, Manary MJ, Trehan I, Mkakosya R, Cheng J, Kau AL, Rich SS, Concannon P, Mychaleckyj JC, Liu J, Houpt E, Li J V, Holmes E, Nicholson J, Knights D, Ursell LK, Knight R, Gordon JI.} 2013. Gut microbiomes of Malawian twin pairs discordant for kwashiorkor. Science {\bf 339}:548–54.
\bibitem{LEA} {\bf Faith JJ, Guruge JL, Charbonneau M, Subramanian S, Seedorf H, Goodman AL, Clemente JC, Knight R, Heath AC, Leibel RL, Rosenbaum M, Gordon JI.} 2013. The long-term stability of the human gut microbiota. Science {\bf 341}:1237439.
\bibitem{hostlife} {\bf David LA, Materna AC, Friedman J, Campos-Baptista MI, Blackburn MC, Perrotta A, Erdman SE, Alm EJ.} 2014. Host lifestyle affects human microbiota on daily timescales. Genome Biol {\bf 15}:R89.
\end{thebibliography}
\end{document}
